Generating Random Logic Programs Using Constraint Programming

### 2020-01-17 paper/paper.tex
An example for predicate independence algorithms

\documentclass[runningheads]{llncs}
\usepackage{graphicx}
%\usepackage[UKenglish]{babel}
%\usepackage[UKenglish]{isodate}
%\usepackage[utf8]{inputenc}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage[capitalise]{cleveref}
\usepackage{tikz}
\usepackage{mathrsfs}
\usepackage[nounderscore]{syntax}

@@

@@

% TODO: example of edge positivity/negativity
\begin{example} \label{ex:program}
  Consider the following program:
  \begin{align}
    \mathsf{sibling}(X, Y) &\gets \mathsf{parent}(X, Z) \land \mathsf{parent}(Y, Z), \nonumber \\
    \mathsf{father}(X, Y) &\gets \mathsf{parent}(X, Y) \land \neg\mathsf{mother}(X, Y). \label{eq:example_clause}
  \end{align}
  Its predicate dependency graph is in \cref{fig:predicate_dependencies}.
  Because of the negation in \cref{eq:example_clause} (as seen in
  \cref{fig:example_tree2}), the edge from $\mathsf{mother}$ to
  $\mathsf{father}$ is negative, while the other two edges are positive.

@@

\begin{definition}[Adjacency matrix representation]
  An $|\predicates{}| \times |\predicates{}|$ adjacency matrix $\mathbf{A}$ with
  $\{ 0, 1 \}$ as its domain is defined by
  \begin{align*}
    \mathbf{A}[i][j] = 0 \iff &\nexists k \in \{ 0, \dots, \maxNumClauses{} \}: \variable{headsOfClauses}[k].\variable{predicate} = j \text{ and } \\
    &i \in \{ a.\variable{name} \mid a \in \variable{bodiesOfClauses}[k].\variable{treeValues} \}.
  \end{align*}
\end{definition}

@@

\begin{algorithm}[p]
  \SetKwFunction{getDependencies}{getDependencies}
  \SetKwFunction{isDetermined}{isDetermined}
  \SetKwData{predicate}{predicate}
  \KwData{predicates $p_1$, $p_2$}
  $D \gets \{ (d_1, d_2) \in \getDependencies{$p_1$, \logical{true}} \times
  \getDependencies{$p_2$, \logical{true}} \mid d_1.\predicate = d_2.\predicate
  \}$\;
  \lIf{$D = \emptyset$}{\Return{\textsc{true}}}
  \lIf{$\exists (\Determined \_, \Determined \_) \in D$}{\Return{\textsc{false}}}
  \Return{\textsc{undefined}}\;
  \caption{Entailment check for predicate independence}
\end{algorithm}

@@

>>>>

\documentclass[runningheads]{llncs}
\usepackage{graphicx}
%\usepackage[UKenglish]{babel}
%\usepackage[UKenglish]{isodate}
%\usepackage[utf8]{inputenc}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage[capitalise]{cleveref}
\usepackage{tikz}
\usepackage{mathrsfs}
\usepackage[nounderscore]{syntax}
\usepackage{blkarray}

@@

\renewcommand\fbox{\fcolorbox{red}{white}}

@@

% TODO: example of edge positivity/negativity
% TODO: this is not a complete program (by my definition). Make it clear in the text.
\begin{example} \label{ex:program}
  Consider the following program:
  \begin{align}
    \mathsf{sibling}(X, Y) &\gets \mathsf{parent}(X, Z) \land \mathsf{parent}(Y, Z), \nonumber \\
    \mathsf{father}(X, Y) &\gets \mathsf{parent}(X, Y) \land \neg\mathsf{mother}(X, Y). \label{eq:example_clause}
  \end{align}
  Its predicate dependency graph is in \cref{fig:predicate_dependencies}.
  Because of the negation in \cref{eq:example_clause} (as seen in
  \cref{fig:example_tree2}), the edge from $\mathsf{mother}$ to
  $\mathsf{father}$ is negative, while the other two edges are positive.

@@

\begin{definition}[Adjacency matrix representation] \label{def:adjacency_matrix}
  An $|\predicates{}| \times |\predicates{}|$ adjacency matrix $\mathbf{A}$ with
  $\{ 0, 1 \}$ as its domain is defined by
  \begin{align*}
    \mathbf{A}[i][j] = 0 \iff &\nexists k \in \{ 0, \dots, \maxNumClauses{} \}: \variable{headsOfClauses}[k].\variable{predicate} = j \text{ and } \\
    &i \in \{ a.\variable{name} \mid a \in \variable{bodiesOfClauses}[k].\variable{treeValues} \}.
  \end{align*}
\end{definition}

@@

\begin{algorithm}[p]
  \SetKwFunction{getDependencies}{getDependencies}
  \SetKwFunction{isDetermined}{isDetermined}
  \SetKwData{predicate}{predicate}
  \KwData{predicates $p_1$, $p_2$}
  $D \gets \{ (d_1, d_2) \in \getDependencies{$p_1$, \logical{true}} \times
  \getDependencies{$p_2$, \logical{true}} \mid d_1.\predicate = d_2.\predicate
  \}$\;
  \lIf{$D = \emptyset$}{\Return{\textsc{true}}}
  \lIf{$\exists (\Determined \_, \Determined \_) \in D$}{\Return{\textsc{false}}}
  \Return{\textsc{undefined}}\;
  \caption{Entailment check for predicate independence}
  \label{alg:independence_entailment}
\end{algorithm}

@@

\begin{figure}[t]
  \centering
\end{figure}

\begin{figure}[t]
  \centering
  \begin{minipage}{.49\textwidth}
    \centering
    \[
      \begin{blockarray}{ccccccc}
        \begin{block}{ccc(cccc)}
          \mathsf{father} & & & 0 & 0 & 0 & 0 \\
          \mathsf{mother} & & & 1 & 0 & 0 & 0 \\
          \mathsf{parent} & & & 1 & \fbox{\{ 0, 1 \}} & \{ 0, 1 \} & \{ 0, 1 \} \\
          \mathsf{sibling} & & & 0 & 0 & 0 & 0 \\
        \end{block}
      \end{blockarray}
    \]
    \caption{The adjacency matrix defined using \cref{def:adjacency_matrix} for
      \cref{example:independence}}
    \label{fig:dependencies_matrix}
  \end{minipage}
  \begin{minipage}{.5\textwidth}
    \centering
    \begin{tikzpicture}
      \node[draw] (parent) at (0, 0.5) {$\mathsf{parent}$};
      \node[draw] (mother) at (0, -0.5) {$\mathsf{mother}$};
      \node[draw] (sibling) at (2, 0.5) {$\mathsf{sibling}$};
      \node[draw] (father) at (2, -0.5) {$\mathsf{father}$};
      \draw[-{Stealth}] (parent) edge (father);
      \draw[-{Stealth}] (mother) edge (father);
      \draw[-{Stealth},dashed,darkgray] (parent) edge (sibling);
      \draw[-{Stealth},dashed,darkgray] (parent) edge (mother);
    \end{tikzpicture}
    \caption{The predicate dependency graph that corresponds to
      \cref{fig:dependencies_matrix}. Dashed edges are undetermined---they may
      or may not exist.}
    \label{fig:dependencies2}
  \end{minipage}
\end{figure}
% TODO: make sure my definitions and examples are consistent with loops

\begin{example} \label{example:independence}
  Consider this partially determined (fragment of a) program:
  \begin{align*}
    \Box(X, Y) &\gets \mathsf{parent}(X, Z) \land \mathsf{parent}(Y, Z),\\
    \mathsf{father}(X, Y) &\gets \mathsf{parent}(X, Y) \land \neg\mathsf{mother}(X, Y)
  \end{align*}
  where $\Box$ indicates an unknown predicate with domain
  \[
    D = \{ \mathsf{father}, \mathsf{mother}, \mathsf{parent}, \mathsf{sibling}
    \}.
  \]
  The predicate dependency graph (without positivity/negativity) defined by
  \cref{def:adjacency_matrix} is represented in
  \cref{fig:dependencies_matrix,fig:dependencies2}.

  Suppose we have a constraint that $\mathsf{mother}$ and $\mathsf{parent}$ must
  be independent. The lists of potential dependencies for both predicates are:
  \begin{align*}
    D_{\mathsf{mother}} &= \{ \Determined \mathsf{mother}, \AlmostDetermined(\mathsf{parent}, \mathsf{parent}, \mathsf{mother}) \}, \\
    D_{\mathsf{parent}} &= \{ \Determined \mathsf{parent} \}.
  \end{align*}
  An entailment check at this stage would produce \textsc{undefined}, but
  propagation replaces the boxed value in \cref{fig:dependencies_matrix} with
  zero, eliminating the potential edge from $\mathsf{parent}$ to
  $\mathsf{mother}$. This also eliminates $\mathsf{mother}$ from $D$, and,
  although some undetermined variables remain, this is enough to make
  \cref{alg:independence_entailment} return \textsc{true}.
\end{example}
